Solution of Differential Algebraic Equations with Manifold Constraints on the Latent Space of an Autoencoder

### 2019-06-10 paper/paper.tex
edited the paper

\setcounter{secnumdepth}{0}

@@

@@

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e& = \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}
(In this paper, we use use $\partial_t$, $\mathrm{d}/\mathrm{d}t$,
and $\dot{()}$ to denote differentiation with respect to time
interchangably depending on context.
The complixity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

@@

We redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. We can cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
where $u=\{T,p,\rho,h\}$. We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \leftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
This maps directly to the original experimental procedure for
determining a constraint equation.

@@

The central idea of this work is to obtain two new variables solve for
those directly,

\begin{align}
\text{Solve for}\, u(t)\, \text{and}\, v(t)\, \text{such that:}\\
\partial_t \rho(u,v) & = \nabla \cdot \mathbf{k} \nabla p(u,v) + r \\
\partial_t \rho e(u,v) & = \nabla \cdot \mathbf{k'}\nabla T(u,v) + s
\end{align}

where \(u\) and \(v\) have no physical meaning other than simply being a
parameterization of the equation of state constraint. We want to select
\(u\) and \(v\) to require no auxilary phase index to define the system
and no additional logic in the code. The sharp kinks and possible
discontinuities can be a part of the functions
\(\rho(u,v), p(u,v), e(u,v)\) and \(T(u,v)\).

Can we automatically discover this parameterization? (For a such a
simple equation, we do not think the proposed methodology will be
better, but the pendulum will instead be our \emph{unit test}.) Our
proposed method is motivated by the problems of multiphase reaction and
transport, where the underlying constraints are the ill-behaved material
properties which are difficult to describe. We change our perspective on
these problems in the next section.

@@

Instead of the painstaking work of parameterizing phase boundaries and
determining curve fits by hand, we rephrase the problem of representing
the equation of state by learning and autoencoder. We have an encoding
phase, \(E(\rho,p,e,T; a)\) and a decoding phase, \(D(u,v; b)\) that
forms an identity function with a compressed subspace:

\begin{equation}
\left\{ \rho, p, e, T \right\} \rightarrow  E \rightarrow \left\{ u,v \right\} \rightarrow D \rightarrow \left\{ \rho, p, e, T \right\}
\end{equation}

The autoencoder is solved for by optimizing its parameters using the
goal

@@

To form the hessian and gradient for this step, we sum over all of the
training data. (The contractive penalty term drops out because it does
not depend on \(W^{dec}\).)

@@

The properties \(p,T\) are the usual prefered choice for primary
variables, but as discussed above, the equations of state are not
analytic for any choice. (Indeed, this idea first spawned when the
author was thinking about systematically rewriting equations in
\(\rho,h\).) Ideally, this system would be fully automated, but due to
the nonlinearilty of the autoencoder problem there are many "bad"
options the trainer can fall into. Multiple starts are possibly needed
to get a good solution when starting from purely randomized newtork
parameters. Global optimization is truly a necessity when trying to
optimize networks with so few parameters. We incorporate a little bit of
human intuition and preference by trying three options to start off the
autoencoder: 1) fully random, 2) \(p,T\) prefered, 3) \(\rho,h\)
prefered. This is implemented by added a bias to the initialization of
the encoder coefficients, e.g. for \(p,T\),
\begin{equation}
 W^{init}_{enc} = \left[\begin{array}{ccccc}
1 & 0 & 0 & 0 & ... \\
0 & 1 & 0 & 0 & ...
                        \end{array}\right]+[\sigma]
\end{equation}
where \([\sigma]\) is the random initialization and \(…\) denotes
padding the coefficients to the nonlinear terms by 0s. Recall that the
values for \(s\) have already been rescaled to be around 0 with a range
of 1. The random initialization part is decided by
\(\sigma \sim \mathcal{N}(0,0.1)\) .

@@

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/slides/classifier_network.png}
\caption{The architecture of the classifier network designed for
  unsupervised labeling of phases.}
\end{figure}

@@

With the trained model, we solve for \(q_1(t)\) and \(q_2(t)\) such that:

\begin{align}
\partial_t \rho(q_1,q_2) & = \nabla \cdot \mathbf{k} \nabla p(q_1,q_2) + r \\
\partial_t \rho h(q_1,q_2)-p & = \nabla \cdot \mathbf{k'}\nabla T(q_1,q_2) + s
\end{align}

where \(\rho(q_1,q_2)\) etc. are the back of an autoencoder:

\begin{equation}
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\} \rightarrow  E \rightarrow 
\left\{ \begin{array}{c} q_1\\q_2 \end{array} \right\}\rightarrow D \rightarrow 
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\}
\end{equation}

@@

First, we had a constrained DAE:

\begin{align}
\frac{\mathrm{d}}{\mathrm{d}t} m(T,p,\rho,h) &= r(T,p,\rho,h)\\
eos(T,p,\rho,h) &= 0
\end{align}

with no good representation of \(eos\) everywhere.

@@

We can parameterize this equation with any time stepping scheme, .e.g. a
Runge-Kutta scheme. For subsurface flow, backward Euler is popular due
to its simplicity and L-stability, which we follow since the topic of
this paper is not exploring new time stepping schemes. This yields the
discretizated algebraic equation for \(q_i\) at \(t+\Delta t\) given the
value \(q^0\) at \(t\),

\begin{equation}
0 = R(q^i) = m(q^{i}) - m(q^{0}) - \Delta t \, r(q^i)
\end{equation}

where we have called the quation \(R(q)\). By design \(R(q)\) is smooth
enough to solve with Newton's method without any statemachine logic:

\begin{equation}
\frac{\partial R}{\partial q} \Delta q^{k+1} = R(q^k)
\end{equation}
This equation is deceptively simple: the function \(R(q)\) has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations XX, XX, and XX require only a few lines
of human-written code.

@@

Since the original manifold is 2D, the user only needs to specify two
initial values. In a multiphase solver such as TOUGH, the state needs
to be specified with the two primary variables plus the state tag: 
\begin{verbatim}
9e5 20 Aqu
0.5 20 AqG
\end{verbatim}
Note that the first primary variable means pressure in the first line
for a pure-liquid state,
but saturation in the second line for the liquid-gas equilibrium.
This makes sense inuitively, as this is how humans are taught, but
still leads to bugs and occasionally requires looking up values in
textbooks or the manual. (Is it $S_{gas}$ or $S_{aqu}$ on the second
line?) A fully fledged solver needs to check the user specified
initial conditions for erroneous inputs.

@@

Note that the encoder only necessarily works when all four values are
correct. Thus, it is necessary to actually solve the nonlinear problem
on the decoder:
\[D(q^0)_i = s^0_i \quad \text{for specified}\; i\]
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The above equation is solved
with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\[K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for specified}\; i, \text{for}\;j=1,2\]
and the following system of equations is solved for an update to $q$,
\[K\Delta q = s^0_i-D(q)_i\]
Updating $q$ is repeated until convergence.

The initial condition problem is not gaurunteed to converge, however,
and often does not.
For example, if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will fail.
To address this issue, we perform multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then pick the
\(q\) that is closest to the two specified components. The multiple
start points are specified at EOS development time to form a set of
values that represent middle points for each phase,
\begin{equation}
\left\{s_{guess}\right\} = \left\{ \left(\begin{array}{c}
\bar{T}_{gas}\\
\bar{p}_{gas}\\
\bar{\rho}_{gas}\\
\bar{h}_{gas}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{liq}\\
\bar{p}_{liq}\\
\bar{\rho}_{liq}\\
\bar{h}_{liq}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{ice}\\
\bar{p}_{ice}\\
\bar{\rho}_{ice}\\
\bar{h}_{ice}
\end{array}\right)... \right\}
  \end{equation}
The multistart method could be specified without any prior knowledge
with random points, but this is a sufficiently robust method for
now. Each of the possible guess points is used to fill in the two
unspecified values as guesses and then passed into the encoder.
E.g. if $p$ and $T$ were specified, each possible $\bar{\rho}$ and
$\bar{h}$ combination from the above sets are filled in to provide
multiple guesses for $q$,
\begin{equation}
s^0 = \left(\begin{array}{c}
T^*(0)\\
p^*(0)\\
\bar{\rho}_{guess}\\
\bar{h}_{guess}
\end{array}\right);\quad q^0_{guess}=E\left(s^0\right)
 \end{equation}
 These sets of $q^0_{guess}$ are then iterated upon using the Newton's
 method procedure. We further add
 under-relaxation to the iteration,
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under relaxation parameter usually set to $\alpha \approx
0.1$.

@@

The codebase is released open source at
\url{https://github.com/afqueiruga/latentsim}. Helper methods that the
author reuses in various TensorFlow-based projects are factored out in
the dully-named \url{https://github.com/afqueiruga/afqstensorflowutils}.
The implementation of the empirical equations of state for water are
factored out at
\href{https://github.com/afqueiruga/equations_of_state}{https://github.com/afqueiruga/equations\emph{of}state}.

@@

\hypertarget{header-n3418}{%
\subsection{Benchmarks}\label{header-n3418}}

The simulator was benchmarke against the long standing TOUGH+ multiphase
multicomponent

@@

@@

Technique ssuch as L1 regularization would be needed to search for low
cost models that also satisfy our desire for Occam's razor
simplifications. (The IAPWS relations used a prelimary form of this
type of simplification, but are not optimal in the author's opinion.)

>>>>

%\setcounter{secnumdepth}{0}

@@

Our
proposed method is motivated by the problems of multiphase reaction and
transport, where the underlying constraints are the ill-behaved material
properties which are difficult to describe.

@@

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e =& \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}
(In this paper, we use use $\partial_t$, $\mathrm{d}/\mathrm{d}t$,
and $\dot{()}$ to denote differentiation with respect to time
interchangably depending on context.
The complixity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

@@

We redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. We can cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
where, for the equation above, the variable is $u=\{T,p,\rho,h\}$. We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \longleftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
This maps directly to the original experimental procedure for
determining a constraint equation.

@@

The central idea of this work is to obtain two new variables solve for
those directly which have no physical meaning other than simply being a
parameterization of the equation of state constraint. We want to select a
\(q_1\) and \(q_2\) to require no auxilary phase index to define the system
and no additional logic in the code. The sharp kinks and possible
discontinuities can be a part of the functions
\(\rho(q_1, q_2), p(q_1, q_2), h(q_1, q_2)\) and \(T(q_1, q_2)\).

@@

Instead of the painstaking work of parameterizing phase boundaries and
determining curve fits by hand, we rephrase the problem of representing
the equation of state by learning and autoencoder. We have an encoding
phase, \(E(\rho,p,e,T; a)\) and a decoding phase, \(D(u,v; b)\) that
forms an identity function with a compressed subspace:
\begin{equation}
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\} \rightarrow  E \rightarrow 
\left\{ \begin{array}{c} q_1\\q_2 \end{array} \right\}\rightarrow D \rightarrow 
\left\{ \begin{array}{c}
T\\ p\\ \rho\\ h
\end{array}\right\}
\end{equation}
The autoencoder is solved for by optimizing its parameters using the
goal

@@

To form the hessian and gradient for this step, we sum over all of the
training data. (The contractive penalty term drops out of the this step because it does
not depend on \(W^{dec}\).)

@@

The properties \(p,T\) are the usual prefered choice for primary
variables, but as discussed above, the equations of state are not
analytic for any choice. (Indeed, this idea first spawned when the
author was thinking about systematically rewriting equations in
\(\rho,h\).) Ideally, this system would be fully automated, but due to
the nonlinearilty of the autoencoder problem there are many poor local minima.
Multiple starts are possibly needed
to get a good solution when starting from purely randomized newtork
parameters. Global optimization is truly a necessity when trying to
optimize networks with so few parameters. We incorporate a little bit of
human intuition and preference by trying three options to start off the
autoencoder: 1) fully random, 2) \(p,T\) prefered, 3) \(\rho,h\)
prefered. This is implemented by added a bias to the initialization of
the encoder coefficients, e.g. for \(p,T\),
\begin{equation}
 W^{init}_{enc} = \left[\begin{array}{ccccc}
1 & 0 & 0 & 0 & ... \\
0 & 1 & 0 & 0 & ...
                        \end{array}\right]+[\sigma]
\end{equation}
where \([\sigma]\) is the random initialization and \(…\) denotes
padding the coefficients to the nonlinear terms by 0s. Recall that the
values for \(s\) have already been rescaled to be around 0 with a range
of 1. The random initialization part is decided by
\(\sigma \sim \mathcal{N}(0,0.1)\) .

@@

The architecture, illustrated in Figure \ref{fig:classifyingnetwork}, has the form:
\begin{align}
  q &= W_{enc} x^n + b_{enc}\\
  f &= W_{dec} q^m + b_{dec} \\
  h &= \sigma\left(W_{bound}q^m+b_{bound}\right) \\
  p &= \text{softmax}\left(W_{phase}q^m+b_{bound[phase}\right) \\
  \bar{x}&= f * p
\end{align}

\begin{figure}
\centering
\includegraphics{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/slides/classifier_network.png}
\caption{\label{fig:classifyingnetwork}The architecture of the classifier network designed for
  unsupervised labeling of phases.}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=3in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/slides/classifier_good.jpg}
  \caption{\label{fig:classified}Result of the unsupervised phase classifier. Compare this plot to the original labeled dataset Figure \ref{fig:pTrho}.}
\end{figure}
Because the identity of the phases,
For comparison purposes, we group together liquid-gas-supercritical as one contiguous label.

@@

With the trained model, we solve for \(q_1(t)\) and \(q_2(t)\) such that:
\begin{align}
\partial_t \rho(q_1,q_2) & = \nabla \cdot \mathbf{k} \nabla p(q_1,q_2) + r \\
\partial_t \rho h(q_1,q_2)-p & = \nabla \cdot \mathbf{k'}\nabla T(q_1,q_2) + s
\end{align}
where \(\rho(q_1,q_2)\) etc. are the back of an autoencoder:

@@

First, we had a constrained DAE:
\begin{align}
\frac{\mathrm{d}}{\mathrm{d}t} m(T,p,\rho,h) &= r(T,p,\rho,h)\\
eos(T,p,\rho,h) &= 0
\end{align}
where $m=\{\rho,\rho h -p\}$ and $r$ is the right hand side, with no good representation of \(eos\) everywhere.

We obtain four new functions with two unknowns. We're actually solving the manifold of Figure X.
\begin{figure}
  \centering
  \includegraphics{../slides/water_four_surfaces_plot.png}
  \caption{\label{fig:foursurfaces}The four 2D surfaces that make up the new parameterization. The learned surface captures the sharp kinks and vertical fronts in Figure \ref{fig:classifyingresult}, but is ananalytic and smooth function of the two learned unknowns.}
  \end{figure}

@@

We can parameterize this equation with any time stepping scheme, .e.g. a
Runge-Kutta scheme. For subsurface flow, backward Euler is popular due
to its simplicity and L-stability, which we follow since the topic of
this paper is not exploring new time stepping schemes. This yields the
discretizated algebraic equation for \(q_i\) and \(t+\Delta t\) given the
value \(q^0\) at \(t\),

\begin{equation}
0 = R(q^i) = m(q^{i}) - m(q^{0}) - \Delta t \, r(q^i)
\end{equation}
where we have called the equation \(R(q)\). By design \(R(q)\) is smooth
enough to solve with Newton's method without any statemachine logic:
\begin{equation}
\frac{\partial R}{\partial q} \Delta q^{k+1} = - R(q^k)
\end{equation}
with the update
\begin{equation}
q^{k+1} = q^k + \alpha\Delta q^{k+1}
\end{equation}
where $\alpha$ is an under relaxation parameter.

This equation is deceptively simple: the function \(R(q)\) has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations XX, XX, and XX requires only a few lines
of human-written code.

@@

Since the original manifold is 2D, the user only needs to specify two
initial values. In a multiphase solver such as TOUGH, the state needs
to be specified with the two primary variables plus the state tag (sans input file formatting): 
\begin{verbatim}
    9e5 20 Aqu
    0.5 20 AqG
\end{verbatim}
Note that the first primary variable means pressure in the first line
for a pure-liquid state,
but saturation in the second line for the liquid-gas equilibrium.
This makes sense inuitively, as this is how humans are taught, but
still leads to bugs and occasionally requires looking up values in
textbooks or the manual. (Is it $S_{gas}$ or $S_{aqu}$ on the second
line?) A fully fledged solver needs to check the user specified
initial conditions for erroneous inputs.

@@

Note that the encoder only necessarily works when all four values are
correct. Thus, it is necessary to actually solve the nonlinear problem
on the decoder:
\[D(q^0)_i = s^0_i \quad \text{for specified}\; i\]
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The above equation is solved
with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\[K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for specified}\; i, \text{for}\;j=1,2\]
and the following system of equations is solved for an update to $q$,
\[K_{ij}\Delta q_j = s^0_i-D(q)_i\]
The iteration u $q$ is repeated until convergence, with under-relaxation used again to update
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under-relaxation parameter usually set to $\alpha \approx
0.1$.
The initial condition problem is not gaurunteed to converge, however,
and often does not.
For example, if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will fail.
To address this issue, we perform multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then pick the
\(q\) that is closest to the two specified components. The multiple
start points are specified at EOS development time to form a set of
values that represent middle points for each phase,
\begin{equation}
\left\{s_{guess}\right\} = \left\{ \left(\begin{array}{c}
\bar{T}_{gas}\\
\bar{p}_{gas}\\
\bar{\rho}_{gas}\\
\bar{h}_{gas}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{liq}\\
\bar{p}_{liq}\\
\bar{\rho}_{liq}\\
\bar{h}_{liq}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{ice}\\
\bar{p}_{ice}\\
\bar{\rho}_{ice}\\
\bar{h}_{ice}
\end{array}\right)... \right\}
  \end{equation}
The multistart method could be specified without any prior knowledge
with random points, but this is a sufficiently robust method for
now. Each of the possible guess points is used to fill in the two
unspecified values as guesses and then passed into the encoder.
E.g. if $p$ and $T$ were specified, each possible $\bar{\rho}$ and
$\bar{h}$ combination from the above sets are filled in to provide
multiple guesses for $q$,
\begin{equation}
s^0 = \left(\begin{array}{c}
T^*(0)\\
p^*(0)\\
\bar{\rho}_{guess}\\
\bar{h}_{guess}
\end{array}\right);\quad q^0_{guess}=E\left(s^0\right)
 \end{equation}
 These sets of $q^0_{guess}$ are then iterated upon using the Newton's
 method procedure.

@@

The codebase is released open source at
\url{https://github.com/afqueiruga/latentsim}. Helper methods that the
author reuses in various TensorFlow-based projects are factored out in
the dully-named \url{https://github.com/afqueiruga/afqstensorflowutils}.
The implementation of the empirical equations of state for water are
factored out at
\href{https://github.com/afqueiruga/equations_of_state}{https://github.com/afqueiruga/equations\_of\_state}.

@@

\hypertarget{header-n3418}{%
\subsection{Validation and Testing}\label{header-n3418}}

The verification problems are listed in Table \ref{tab:problems}.
\begin{table}
  \caption{\label{tag:problems}Enumeration of the problems used to test and evaluate each fo the generaed architectures}
\begin{tabular}{l|c|c|c}
  Name & start & path & end\\
\hline
\end{tabular}
\end{table}
The simulator was benchmarke against the long standing TOUGH+ multiphase
multicomponent

@@

We can also think about this work as offline preconditioning.

In TOUGH, for one instance, all of the calculations are performmed in SI units because it is simply too much labor to normalize everything! Adverse effects to iterative algorithms and floating point error propogation occur, especially because numerical differentiation is used.
The learned primary variables are normalized to a unit range around zero, and the scaling matrix is explicitly in one step. The scaling matrix could be a way to systematically nondimensionalize the entire problem.

@@

Techniquessuch as L1 regularization would be needed to search for low
cost models that also satisfy our desire for Occam's razor
simplifications. (The IAPWS relations used a prelimary form of this
type of simplification, but are not optimal in the author's opinion.)
